---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ggwordcloud

`ggwordcloud` provides a word cloud text geom for `ggplot2`.  The placement algorithm implemented in C++ is close to the one used in `wordcloud2.js` and thus aims to be a replacement of `wordcloud2` that produces `ggplot2` type plot instead of an html widget. Note that the current version does not provide the shape and mask possibility of `wordcloud2`. The algorithm of `wordcloud` is similar but `ggwordcloud` allows arbitrary rotations of the words.

## Installation

You can install the released version of ggwordcloud from [CRAN](https://CRAN.R-project.org) with:
```{r, eval=FALSE}
install.packages("ggwordcloud")
```
or the development version from the github repository
```{r, eval=FALSE}
devtools::install_github("lepennec/ggwordcloud")
```

# A simple word cloud example

As an example, we will use the `mtcars` dataset 
```{r}
library(ggwordcloud)
dat <- mtcars
dat$name <- row.names(mtcars)
dat$size <- dat$mpg
dat$size[c(1,4)] <- dat$size[c(1,4)] + 100
set.seed(42)
ggplot(data = dat, aes(label = name, size = size)) + geom_text_wordcloud() +
  theme_minimal()
```

More examples are available in the vignette.

