---
title: "ggwordcloud: a wordcloud geom for ggplot2"
author: "E. Le Pennec"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggwordcloud: a wordcloud geom for ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`ggwordcloud` provides a wordcloud text geom for `ggplot2`.  The placement algorithm implemented in C++ is close to the one used in `wordcloud2.js` and thus aims to be a replacement of `wordcloud2` that produces `ggplot2` type plot instead of an html widget. Note that the current version does not provide the shape and mask possibility of `wordcloud2`. Even if the algorithm of `wordcloud` is similar, `ggwordcloud` is much faster and allows arbitrary rotations of the words.

This vignette is meant as a quick tour of its possibility.

# Package installation

The package can be installed from CRAN by
```{r, eval=FALSE}
install.packages("ggwordcloud")
```
or the development version from the github repository
```{r, eval=FALSE}
devtools::install_github("lepennec/ggwordcloud")
```

# Wordcloud

As an example, we will use the `mtcars` dataset 
```{r}
dat <- mtcars
dat$name <- row.names(mtcars)
```

Let's load the package and set the random seed.
```{r}
library(ggwordcloud)
set.seed(42)
```

A basic wordcloud can be obtained by:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name)) + geom_text_wordcloud() + theme_minimal()
```
We have used a minimal theme to display only the words.


Because there is some randomness in the algorithm, the same command can yield a different result:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name)) + geom_text_wordcloud() + theme_minimal()
```

# Wordcloud and text size

We will use the `mpg` variable to define a size and modify two cars so that they are much larger than the other ones:
```{r}
dat$size <- dat$mpg
dat$size[c(1, 4)] <- dat$size[c(1, 4)] + 100
```

We can now add this new variable to the aesthetic:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size)) + geom_text_wordcloud() +
  theme_minimal()
```

In order to obtain a better picture, one can play with the size scale:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size)) + geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

# Wordcloud and text area

By default, the size of the font depends linearly from the square root of the size aesthetic rather than the raw size aesthetic. This is a natural choice for a shape as the area of the shape will be proportional to the raw size aesthetic but not necessarily for texts with different lengths. In `ggwordcloud2`, there is an option, `area_corr` to scale the texts so that their area is a function of the raw size aesthetic when used in combination with `scale_size_area`:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size)) +
  geom_text_wordcloud(area_corr = TRUE) +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

By default, the area is proportional to the raw size aesthetic raised to the power $1/.7$ in order to match the human area perception. To obtain an area proportional to the raw size aesthetic, it suffices to set the `area_corr_power` to $1$:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size)) +
  geom_text_wordcloud(area_corr = TRUE, area_corr_power = 1) +
  scale_size_area(max_size = 12) +
  theme_minimal()
```


# Wordcloud with too many words

Note that words that cannot be placed due to a lack of space are by default displayed at their original position:
```{r, fig.width=7.5, fig.height=4}
set.seed(42)
ggplot(data = dat, aes(label = name, size = size)) + geom_text_wordcloud() +
  scale_size_area(max_size = 20) +
  theme_minimal()
```
It is up to the user to avoid this issue by either removing some words or changing the size scale.

One can also chose to remove those words using the `rm_outside` option:
```{r, fig.width=7.5, fig.height=4}
set.seed(42)
ggplot(data = dat, aes(label = name, size = size)) +
  geom_text_wordcloud(rm_outside = TRUE) +
  scale_size_area(max_size = 20) +
  theme_minimal()
```



# Wordcloud and rotation

Let's start by creating a rotation angle of 90 for 40 \% of the words:
```{r}
dat$rot <- 90 * (runif(nrow(dat)) > .6)
```
We can use this variable in the aesthetic:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

`ggwordcloud` allows arbitrary rotations:
```{r}
dat$rot <- (-90 + 180 * runif(nrow(dat))) * (runif(nrow(dat)) > .2)
```

```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

# Wordcloud and eccentricity

The `ggwordcloud` algorithm moves the text around a spiral until it finds a free place for it.
This spiral has by default a vertical eccentricity of .65,
so that the spiral is 1/.65 wider than taller.

```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

This can be changed using the `eccentricity` parameter:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud(eccentricity = 1) +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud(eccentricity = .35) +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

# Advanced wordcloud

`geom_text_wordcloud` is compatible with the facet system of `ggplot2`:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(label = name, size = size, angle = rot)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  facet_wrap(~am) +
  theme_minimal()
```


One can also specify an original position for each label that what will be used as the starting point of the spiral algorithm for this label:
```{r, fig.width=7.5, fig.height=5}
ggplot(data = dat, aes(
  x = factor(am), label = name, size = size, angle = rot,
  color = factor(am)
)) +
  geom_text_wordcloud() +
  scale_size_area(max_size = 12) +
  theme_minimal()
```

# `ggwordcloud` as an approximate replacement for `wordcloud` and `wordcloud2`

`ggwordcloud` and `ggwordcloud2` are two approximate replacement for respectively `wordcloud` and `wordcloud2`. They provide a similar syntax than the original functions and yields similar wordclouds, but not all the options of the original functions are implemented. Note that both use a font size proportional to the raw size aesthetic. This is implemented in `ggwordcloud` by the use of the `scale_radius` size scale.

```{r, fig.width=7.5, fig.height=5}
ggwordcloud(dat$name, dat$size)
```

```{r, fig.width=7.5, fig.height=5}
ggwordcloud2(dat[, c("name", "size")])
```


